---
title: "Analysis of US Consumer Price Index"
author: "Mehedi Hasan Jim"
date: "2025-07-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache =  TRUE)
```

<style> body {text-align:justify;  font-family: "Georgia";}
#header{text-align: center}
.r{background-color: #caf0f8;}
</style>
<br>
<br>

We will use the US Consumer Price Index (CPI) data set from the *tseries* package.

The data set is part of *NelPlo data.*

```{r cars}
data(NelPlo, package = "tseries")
head(cpi)
cpi <- ts(cpi,start = 1860, end = 1988)
```




## Test and Training Set

To understand which forecasting method performs the best it's necessary to evaluate them based on training and testing data. We take the last 8 years for testing data.


```{r}
library(tidyverse)
library(tseries)
library(forecast)
library(urca)
cpi <- ts(cpi,start = 1860, end = 1988)
cpi.train <- window(cpi, end = 1980)
cpi.test <- window(cpi, start = 1981)
head(cpi.test)
```


## ploting the data

It's essential to plot the data to understand it. Plotting the data will help to get idear about the nature of the data and the potential models that can be applied.

```{r}
autoplot(cpi.train, linewidth = 1)+
  labs(tittle = "Consumer Price Index: Training Data",
       y= "",
       caption = "Source: C. R. Nelson and C. I. Plosser (1982),\n Trends and Random Walks in Macroeconomic Time Series. \n Journal of Monetary Economics, 10, 139–162.\n doi:10.1016/0304-3932(82)90012-5.")+
  scale_x_continuous(breaks= c(1860, 1900, 1950, 1988))+
  theme_classic()+
  theme(text = element_text(size = 20, family = "serif"),
        plot.caption = element_text(size = 7))
```



Let's take first difference of the data to see whether the trend/ drift goes away or not.


```{r}

autoplot(diff(cpi.train,2), linewidth = 1)+
  labs(title= expression(Delta^2 * "Consumer Price Index"), y = "",
                 caption = "Source: C. R. Nelson and C. I. Plosser (1982),\n Trends and Random Walks in Macroeconomic Time Series. \n Journal of Monetary Economics, 10, 139–162.\n doi:10.1016/0304-3932(82)90012-5.")+
  scale_x_continuous(breaks= c(1860, 1900, 1950, 1988))+
  theme_classic()+
  theme(text = element_text(size = 20, family = "serif"),
        plot.caption = element_text(size = 7))
```


Seems like differencing is making it trend stationary. Need to ensure wheater the variable at level and/or variable at first difference has unit root or not.
Note: A series with unit root has time variant mean and/or variance.

Now let's explore the unit root, ACF and PACF to understand what will be the potential values of p,d and q in the *ARIMA* model

```{r}
urca::ur.df(cpi.train,type = "drift") |>summary() # alternative tests are kpss, ers, pp, sp, za tests from the urca package

```

The unit root test (Null: The series has unit root.) shows that CPI has unit root at level.

```{r}
urca::ur.df(diff(cpi.train),type = "drift") |>summary() 

```


The test result shows that the first differnce series dos not have unit root.


```{r}
Acf(diff(cpi.train))

```



```{r}
Pacf(diff(cpi.train))

```

Seems like ARIMA(2,1,2) can be a good choice with two significant spikes in each of the plots. Let's see what does *auto.arima()* say.

```{r}

cpi.fit.auto.arima <- auto.arima(cpi.train, seasonal = F)

```
## forecast using auto.arima

```{r}
forecast.auto.arima <- cpi.fit.auto.arima%>% forecast(h = 8)
autoplot(cpi, color = "darkgreen", size =1, series = "CPI")+
  autolayer(forecast.auto.arima, PI =F, color = "darkred", size = 1, series = "Arima Prediction")+
  theme_classic()

```


Notice that the forecasted values are very close to the real values. Remember that the last 8 years were not used in training. Seems a good fit. Let's evaluate the fit.

## ARIMA fit Evaluation

```{r}

accuracy(forecast.auto.arima, cpi.test) %>% round(3)

```


NOt a bad result. Let's fit some other models and compare them.



## Holt's Exponential Smoothing

Since no seasonality is possible for the data under study, we will use a simple Holt's model.


```{r}

cpi.fit.holt <- holt(cpi.train, h = 10)
accuracy(cpi.fit.holt, cpi.test) %>% round(3)

```


```{r}
forecast.holt <- forecast(cpi.fit.holt,h = 8)
autoplot(cpi, color = "darkgreen", size =1, series = "CPI")+
  autolayer(forecast.auto.arima, PI =F, color = "darkred", size = 1, series = "Arima Prediction")+
  autolayer(forecast.holt, size= 1, color = "blue", PI = F)+
  theme_classic()
```
Let's get to the damped Holt's model

```{r}
cpi.fit.holt.damped <- holt(cpi.train, h = 10,damped = T)
accuracy(cpi.fit.holt.damped, cpi.test) %>% round(3)

forecast.holt.damped <- forecast(cpi.fit.holt,h = 8)
autoplot(cpi, color = "darkgreen", size =1, series = "CPI")+
  autolayer(forecast.auto.arima, PI =F, color = "darkred", size = 1, series = "Arima Prediction")+
  autolayer(forecast.holt, size= 1, color = "blue", PI = F)+
  autolayer(forecast.holt.damped, size= 1, linetype = 3, color = "purple", PI = F)+
  theme_classic()
```


The damped model predicted the same as the usual model. So far the ARIMA seems a better fit. 
Note: we could use the ets() function here.


```{r}



```

